---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
# Set matplotlib backend for headless rendering
Sys.setenv(MPLBACKEND = "Agg")
```

# rminte

<!-- badges: start -->
[![R-CMD-check](https://github.com/mrc-ide/rminte/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mrc-ide/rminte/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/mrc-ide/rminte/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/mrc-ide/rminte/actions/workflows/pkgdown.yaml)
[![Codecov test coverage](https://codecov.io/gh/mrc-ide/rminte/branch/main/graph/badge.svg)](https://app.codecov.io/gh/mrc-ide/rminte?branch=main)
<!-- badges: end -->

An R interface to the Python [MINTe (Malaria Intervention Emulator)](https://github.com/CosmoNaught/MINTe-python) package. This package provides neural network-based malaria scenario predictions for evaluating intervention strategies including ITNs (Insecticide-Treated Nets), IRS (Indoor Residual Spraying), and LSM (Larval Source Management).

## Documentation

📖 **[Full documentation and vignettes](https://mrc-ide.github.io/rminte/)**

## Features

- **Fast Scenario Predictions**: Run thousands of malaria intervention scenarios in seconds using pre-trained LSTM models
- **Multiple Predictors**: Predict both prevalence and clinical cases
- **Flexible Net Types**: Support for various ITN types (pyrethroid-only, PBO, pyrrole, PPF)
- **Model Caching**: Efficient caching of loaded models for faster subsequent runs
- **Native R Plotting**: Built-in ggplot2 visualization functions
- **Web Interface Support**: Lightweight controller for web application integration

## Installation

### Prerequisites

First, install the Python `minte` package:

```bash
pip install minte
```

### Install rminte

Install the R package from GitHub:

```r
# install.packages("devtools")
devtools::install_github("mrc-ide/rminte")
```

Alternatively, install the Python package from within R:

```r
library(rminte)
install_minte()
```

## Quick Start

### Check Installation

```{r check-install}
library(rminte)

# Check if minte is available
minte_available()
```

### Run a Single Scenario

```{r single-scenario}
results <- run_minter_scenarios(
 scenario_tag = "example_scenario",
 res_use = 0.2,
 py_only = 0.3,
 py_pbo = 0.2,
 py_pyrrole = 0.1,
 py_ppf = 0.05,
 prev = 0.55,
 Q0 = 0.92,
 phi = 0.85,
 season = 0,
 irs = 0.4,
 irs_future = 0.4,
 lsm = 0.2,
 routine = 1,
 itn_future = 0.45,
 net_type_future = "py_only"
)

# View results structure
print(results)
```

```{r view-prevalence}
# View prevalence predictions
head(results$prevalence)
```

### Running Multiple Scenarios

```{r multiple-scenarios}
# Define multiple scenarios
n_scenarios <- 5
results <- run_minter_scenarios(
 scenario_tag = c("no_intervention", "irs_only", "lsm_only", 
                  "py_pbo_only", "py_pbo_with_lsm"),
 res_use = rep(0.2, n_scenarios),
 py_only = c(0.3, 0.3, 0.3, 0.0, 0.0),
 py_pbo = c(0.2, 0.2, 0.2, 0.5, 0.5),
 py_pyrrole = rep(0.1, n_scenarios),
 py_ppf = rep(0.05, n_scenarios),
 prev = rep(0.55, n_scenarios),
 Q0 = rep(0.92, n_scenarios),
 phi = rep(0.85, n_scenarios),
 season = rep(0, n_scenarios),
 irs = c(0.0, 0.5, 0.0, 0.0, 0.0),
 irs_future = c(0.0, 0.5, 0.0, 0.0, 0.0),
 lsm = c(0.0, 0.0, 0.3, 0.0, 0.3),
 routine = rep(1, n_scenarios)
)

# Analyze results
library(dplyr)
results$prevalence %>%
 group_by(scenario_tag) %>%
 summarise(mean_prev = mean(prevalence)) %>%
 arrange(mean_prev)
```

### Visualization

```{r plot-prevalence, fig.width=10, fig.height=6}
library(ggplot2)

# Plot prevalence over time
p <- plot_prevalence(results, title = "Malaria Prevalence by Intervention")
print(p)
```

```{r plot-cases, fig.width=10, fig.height=6}
# Plot cases over time
p_cases <- plot_cases(results)
print(p_cases)
```

### Utility Functions

```{r utility-functions}
# Calculate net effectiveness from resistance
result <- calculate_overall_dn0(
 resistance_level = 0.3,
 py_only = 0.4,
 py_pbo = 0.3,
 py_pyrrole = 0.2,
 py_ppf = 0.1
)
cat("Overall dn0:", result$dn0, "\n")
cat("Total ITN use:", result$itn_use, "\n")

# Get available net types
net_types <- available_net_types()
print(net_types)
```

## API Reference

### Main Functions

| Function | Description |
|----------|-------------|
| `run_minter_scenarios()` | Main entry point for running intervention scenarios |
| `run_malaria_emulator()` | Direct emulator interface for scenario predictions |
| `run_mintweb_controller()` | Simplified web interface controller |
| `plot_prevalence()` | Plot prevalence over time (ggplot2) |
| `plot_cases()` | Plot cases over time (ggplot2) |
| `create_scenario_plots_mpl()` | Create visualizations from results (matplotlib) |

### Utility Functions

| Function | Description |
|----------|-------------|
| `calculate_overall_dn0()` | Calculate net effectiveness from resistance and coverage |
| `available_net_types()` | Get list of supported net types |
| `resistance_to_dn0()` | Convert resistance to net effectiveness |
| `preload_all_models()` | Pre-load models into cache |
| `clear_cache()` | Clear model cache |
| `create_scenarios()` | Helper to create scenarios DataFrame |

### Setup Functions

| Function | Description |
|----------|-------------|
| `minte_available()` | Check if minte Python package is available |
| `configure_minte()` | Configure Python environment before initialization |
| `install_minte()` | Install minte Python package (for manual env management) |

## Output Structure

The `run_minter_scenarios()` function returns a list with:

- **prevalence**: Data frame with columns:
    - `index`: Scenario index
    - `timestep`: Time index (14-day periods)
    - `prevalence`: Under-5 prevalence
    - `model_type`: Model used (e.g., "LSTM")
    - `scenario` / `scenario_tag`: Scenario identifier
    - `eir_valid`: Whether EIR is within calibrated range
- **cases**: Data frame with similar structure, containing clinical cases per 1000
- **scenario_meta**: Per-scenario metadata
- **eir_valid**: Overall EIR validity flag
- **benchmarks**: Timing information (if `benchmark = TRUE`)

## Requirements

- R >= 4.0.0
- Python >= 3.8
- reticulate >= 1.41
- minte Python package

Optional:
- ggplot2 (for native R plotting)
- dplyr (for data manipulation examples)

## License

MIT

## Citation

If you use this package, please cite the original MINTe Python package:
- [MINTe-python](https://github.com/CosmoNaught/MINTe-python)

## Related Projects

- [MINTe-python](https://github.com/CosmoNaught/MINTe-python) - Original Python package
- [estiMINT-python](https://github.com/CosmoNaught/estiMINT-python) - EIR estimation package
